---
title: "TN-seq phage"
output: html_notebook
editor_options: 
  chunk_output_type: console
---


```{r}
library(tidyverse)
```

This markdown doc contains the code required to make the figures in the paper on phage gene essentially and socialility called "Sociality changes gene essentiality"


Figure 1 simple model showing where the MOI threshold is for cells to on average have more than 1 phage
```{r}

p_ge_2 <- function(moi) {
  1 - ppois(1, lambda = moi)  # 1 - P(K <= 1)
}

# ---- Solve for MOI where P(K >= 2) = 0.5 ----
# We solve p_ge_2(moi) - 0.5 = 0 over a reasonable interval
threshold <- uniroot(function(m) p_ge_2(m) - 0.5, interval = c(0, 10))$root

# ---- Make a curve over MOI values ----
moi_grid <- seq(0, 6, by = 0.01)

p0   <- dpois(0, lambda = moi_grid)           # P(K = 0)
p1   <- dpois(1, lambda = moi_grid)           # P(K = 1)
pge2 <- 1 - ppois(1, lambda = moi_grid)       # P(K >= 2)

# ---- Plot the diagram ----
plot(moi_grid, pge2, type = "l", lwd = 2,
     xlab = "MOI (mean phage adsorptions per cell)",
     ylab = "Fraction of cells",
     ylim = c(0, 1),
     main = "")

# Add helpful reference curves (optional but informative)
#lines(moi_grid, p0, lwd = 2, lty = 2)  # P(K=0)
#lines(moi_grid, p1, lwd = 2, lty = 3)  # P(K=1)

# Horizontal line at 0.5 (majority threshold)
abline(h = 0.5, lty = 2)

# Vertical line at the threshold MOI
abline(v = threshold, lwd = 2, lty = 2)

# Mark the threshold point on the P(K>=2) curve
points(threshold, 0.5, pch = 19)

# Label it
#text(threshold+1, 0.55,
#     labels = paste0("P(K>=2)=0.5 at MOI ≈ ", round(threshold, 3)),
#     pos = 4)

#legend("bottomright",
#       legend = c("P(K >= 2)", "P(K = 0)", "P(K = 1)"),
       #lty = c(1, 2, 3),
       #lwd = 2,
       #bty = "n")

# Print the threshold value to the console too
cat("MOI threshold where majority have >1 adsorption (P(K>=2)=0.5):",
    threshold, "\n")



```

Figure 2 basic model showing social genes will increase in freq
```{r}

N_infections <- rpois(1e4, 5 * 1) # (number of cells, moi * prop of wt or KO)

hist(N_infections) # yeah this is the same as the infection dynamics MOI code I had from ages ago. 



simulate_simple <- function(
    x_wt = 0.99, # preportion of non mutant at a allele 
    moi = 1, # moi duh
    sigma = 1,        # 0 = private, 1 = social
    n_cells = 100000,
    seed = 1337
) {
  set.seed(seed)
  
  x_ko <- 1 - x_wt # how many knockouts in prop to wt
  
  # Infection counts
  N_wt <- rpois(n_cells, moi * x_wt)
  N_ko <- rpois(n_cells, moi * x_ko)
  K <- N_wt + N_ko
  
  infected <- K > 0
  N_wt <- N_wt[infected]
  N_ko <- N_ko[infected]
  
  # Cases
  wt_only   <- N_wt > 0 & N_ko == 0
  ko_only   <- N_ko > 0 & N_wt == 0
  coinfect  <- N_wt > 0 & N_ko > 0
  
  # WT always produces
  wt_progeny <- sum(wt_only) + sum(coinfect)
  
  # KO produces only if:
  # - coinfected
  # - rescued with probability sigma
  rescued_coinfect <- rbinom(sum(coinfect), 1, sigma)
  ko_progeny <- sum(rescued_coinfect)
  
  total <- wt_progeny + ko_progeny
  
  x_wt_next <- wt_progeny / total
  x_ko_next <- ko_progeny / total
  
  c(WT = x_wt_next, KO = x_ko_next)
}


mois <- c(0.1, 0.5, 1, 2, 5, 10,25)

res_private <- sapply(mois, function(m)
  simulate_simple(x_wt = 0.5, moi = m, sigma = 0)["KO"]
)

res_social <- sapply(mois, function(m)
  simulate_simple(x_wt = 0.5, moi = m, sigma = 1)["KO"]
)

plot(mois, res_private, type="b", log="x", col="red", ylim=c(0,0.6),
     xlab="MOI", ylab="KO frequency after passage")
lines(mois, res_social, type="b", col="blue")
legend("topleft",
       legend=c("Private (σ=0)", "Social (σ=1)"),
       col=c("red","blue"), lty=1)

```


figure 3 version  with random draw of mutants 
```{r}
################################################################################
# TnSeq Phage Simulation 
#
# This code takes 99 genes assigns them a random fitness score, drawn from a bounded uniform distribution
# and a random complementation score (0 being no complementation can happen and 1 being a gene is full complemented in the preseance of another). 
# currently if outputs multiple figures, only p2 is going into the paper I think.
#
################################################################################

library(tidyverse)

# ─────────────────────────────────────────────────────────────────────────────
# 1. GENE PARAMETER TABLE
# ─────────────────────────────────────────────────────────────────────────────

set.seed(42)

n_genes <- 375

gene_params <- tibble(
  gene_id = paste0("gene_", sprintf("%02d", 1:n_genes)),

  # True fitness effect of disruption (0 = lethal, 1 = fully neutral)
  fitness_effect = c(
    runif(n_genes/3, 0.00, 0.05),   # essential genes
    runif(n_genes/3, 0.40, 0.60),   # moderate-effect genes
    runif(n_genes/3, 0.70, 1.00)    # near-neutral genes
  ) %>% sample(),

  # Complementation fraction: how much fitness is rescued when a co-infecting
  # phage supplies a functional copy of this gene
  # 0 = not complementable (cis-acting)
  # 1 = fully complementable (trans or diffusible enzyme)
  complementation = c(
    runif(n_genes/3, 0.00, 0.10),
    runif(n_genes/3, 0.40, 0.80),
    runif(n_genes/3, 0.85, 1.00)
  ) %>% sample()
)

print(gene_params, n = 375) # does this look right?


# ─────────────────────────────────────────────────────────────────────────────
# 2. HELPER: COUNT FUNCTIONAL CO-INFECTORS FOR A GIVEN GENE
# ─────────────────────────────────────────────────────────────────────────────

#' Given a vector of co-infecting phage gene disruptions, count how many
#' are functional (i.e. NOT disrupted) at the focal gene.
#'
#' @param focal_gene      Integer: which gene the focal mutant is disrupted in
#' @param coinfector_genes Integer vector: which gene each co-infector is disrupted in
#' @return Integer: number of co-infectors that can complement focal_gene

count_helpers <- function(focal_gene, coinfector_genes) {
  sum(coinfector_genes != focal_gene)
}


# ─────────────────────────────────────────────────────────────────────────────
# 3. FITNESS OF ONE PHAGE
# ─────────────────────────────────────────────────────────────────────────────

#' Calculate relative phage output for one mutant phage in one cell
#'
#' @param focal_gene      Which gene is disrupted in the focal phage
#' @param n_helpers       Number of co-infectors that are WT at focal_gene
#' @param gene_params     Gene parameter tibble
#' @return Relative phage output (0–1)
# this is bad to name to style params the same but this wont be a package so it is fine.

phage_output <- function(focal_gene, n_helpers, gene_params) {
  gp          <- gene_params[focal_gene, ] #
  intrinsic   <- gp$fitness_effect

  if (n_helpers == 0) {
    return(intrinsic)
  } else {
    # At least one helper present: rescue fitness proportional to complementation
    rescued <- intrinsic + gp$complementation * (1 - intrinsic)
    return(rescued)
  }
}

# to break this function down, if there isn't another phage present then you return the fitness effect
# if there is then you get the else statement which is the rescued call.
# as an example lets say you have mutation that results in a low fitness alone of 0.2 but if theres another phage there then you get a compl score of 0.8. this results in a final fitness effect of 0.68 
#0.2 + 0.8 * (0.8-0.2)



# ─────────────────────────────────────────────────────────────────────────────
# 4. CORE SIMULATION — SATURATED MUTANT LIBRARY
# ─────────────────────────────────────────────────────────────────────────────

#' Simulate one TnSeq experiment with a saturated mutant library
#'
#' Every phage carries an insertion in one randomly chosen gene.
#' Co-infecting phages can complement a focal mutant only if their insertion
#' is in a DIFFERENT gene.
#'
#' @param moi          Mean MOI — average number of phages per bacterium (Poisson)
#' @param n_bacteria   Number of bacteria in the infection
#' @param burst_size   Mean WT burst size (Poisson-distributed output per cell)
#' @param gene_params  Gene parameter tibble
#' @param seed         Random seed
#' @return Tibble with per-gene estimated fitness scores

#so this is just getting the function ready. The main simulation will have a range of MOIS
simulate_tnseq_v1b <- function(moi          = 1,
                                n_bacteria   = 2e5,
                                burst_size   = 50,
                                gene_params  = gene_params,
                                seed         = NULL) {

  if (!is.null(seed)) set.seed(seed)
  n_genes <- nrow(gene_params)

  # Track input and output counts per gene
  input_counts  <- numeric(n_genes)
  output_counts <- numeric(n_genes)

  # Draw number of phages per cell from Poisson(MOI)
  phages_per_cell <- rpois(n_bacteria, lambda = moi)
  phages_per_cell <- phages_per_cell[phages_per_cell > 0]
  n_infected      <- length(phages_per_cell)

  for (cell in seq_len(n_infected)) {
    n_phage <- phages_per_cell[cell]

    # Every phage has an insertion — draw which gene for each phage
    # Uniform across all genes (can be made non-uniform)
    cell_genes <- sample(n_genes, n_phage, replace = TRUE)

    # Each phage is the "focal" mutant in turn; the rest are co-infectors
    for (p in seq_len(n_phage)) {
      focal_gene      <- cell_genes[p]
      coinfector_genes <- cell_genes[-p]

      input_counts[focal_gene] <- input_counts[focal_gene] + 1

      # How many co-infectors are WT at the focal gene?
      n_helpers <- count_helpers(focal_gene, coinfector_genes)

      # Relative fitness
      rel_fitness <- phage_output(focal_gene, n_helpers, gene_params)

      # Stochastic burst proportional to fitness, does this work...?
      actual_output <- rpois(1, lambda = rel_fitness * burst_size)

      output_counts[focal_gene] <- output_counts[focal_gene] + actual_output
    }
  }

  # ── Estimate fitness score ──────────────────────────────────────────────────
  pseudo          <- 1
  log2_ratio      <- log2((output_counts + pseudo) / (input_counts + pseudo))

  # We normalize reads, do we need to do it here? Lets try normalize to median of near-neutral genes
  neutral_idx     <- which(gene_params$fitness_effect > 0.7)
  log2_ratio_norm <- log2_ratio - median(log2_ratio[neutral_idx]) #log2 median normalization 

  tibble(
    gene_id         = gene_params$gene_id,
    true_fitness    = gene_params$fitness_effect,
    complementation = gene_params$complementation,
    input_counts    = input_counts,
    output_counts   = output_counts,
    log2_ratio      = log2_ratio,
    fitness_score   = log2_ratio_norm,
    moi             = moi
  )
}


# ─────────────────────────────────────────────────────────────────────────────
# 5. ANALYTICAL EXPECTATION: P(at least one helper) given MOI and n_genes
# ─────────────────────────────────────────────────────────────────────────────

#' For a focal phage disrupted in gene X, what is the probability that at
#' least one co-infecting phage is WT at gene X?
#'
#' Each co-infector independently has probability (1 - 1/n_genes) of being
#' WT at gene X (i.e. its insertion is somewhere else).
#' The number of co-infectors given the focal phage is already in the cell
#' follows a Poisson(MOI) distribution (approximately).
#'
#' P(at least one helper) = 1 - P(no helpers)
#'   = 1 - sum_{k=0}^{inf} P(k co-infectors) * (1/n_genes)^k
#'   = 1 - exp(-MOI * (1 - 1/n_genes))   [Poisson pgf result]

# below helps us work out the rate at which the same gene is broken with x infecting phages
# maybe interesting maybe not but by definition that means that smaller genomes are more likely to have the same knock out.

p_at_least_one_helper <- function(moi, n_genes) {
  p_same_gene <- 1 / n_genes
  1 - exp(-moi * (1 - p_same_gene))
}

# If intuition isn't clear, lets make a table with more and more genes and ask how the rate of helpers changes in response to phage genome size and moi 
helper_prob_table <- expand_grid(
  moi     = c(0.1, 0.5, 1, 2, 5, 10),
  n_genes = c(10, 20, 40, 80, 160, 320)
) %>%
  mutate(p_helper = p_at_least_one_helper(moi, n_genes))

cat("\nP(at least one helper co-infector) by MOI and genome complexity:\n")
helper_prob_table %>%
  pivot_wider(names_from = n_genes, values_from = p_helper,
              names_prefix = "n_genes=") %>%
  print()

# its not a big effect but as you insrease the number of genes the % incrases that a phage doesnt have the same mutation as you. This is all a bit of a distraction but worth noting. 

# ─────────────────────────────────────────────────────────────────────────────
# 6. RUN SIMULATION ACROSS MOIs
# ─────────────────────────────────────────────────────────────────────────────

# now we can use the simulate_tn-seq function we got earlier. This will take a min or two to run.

moi_values <- c(0.01, 0.1, 0.5, 1, 2, 5, 10)


cat("\nRunning simulations...\n")
results_all <- map_dfr(moi_values, function(m) {
  cat("  MOI =", m, "\n")
  simulate_tnseq_v1b(
    moi         = m,
    n_bacteria  = 2e5,
    burst_size  = 50,
    gene_params = gene_params,
    seed        = 123
  )
})


# ─────────────────────────────────────────────────────────────────────────────
# 7. VISUALIZATIONS
# ─────────────────────────────────────────────────────────────────────────────

results_all <- results_all %>%
  mutate(
    gene_class = case_when(
      true_fitness < 0.1  ~ "Essential",
      true_fitness < 0.6  ~ "Moderate",
      TRUE                ~ "Near-neutral"
    ),
    comp_class = case_when(
      complementation < 0.2 ~ "Low comp.",
      complementation < 0.7 ~ "Moderate comp.",
      TRUE                  ~ "High comp."
    )
  )

# ── Plot 1: Fitness score vs MOI, faceted by gene class ──────────────────────
p1 <- ggplot(results_all,
             aes(x = factor(moi), y = fitness_score,
                 color = gene_class, group = gene_id)) +
  geom_line(alpha = 0.5, linewidth = 0.6) +
  geom_point(alpha = 0.7, size = 1.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  facet_wrap(~gene_class) +
  scale_color_manual(values = c(Essential    = "#d62728",
                                Moderate     = "#ff7f0e",
                                "Near-neutral" = "#2ca02c")) +
  labs(
    title    = "TnSeq Fitness Score vs MOI — Saturated Mutant Library",
    subtitle = "Every phage carries a unique insertion; co-infectors complement only if disrupted in a different gene",
    x        = "MOI",
    y        = "Normalized log2(output / input)",
    color    = "Gene class"
  ) +
  theme_bw(base_size = 12) +
  theme(legend.position = "bottom")

print(p1)
#ggsave("tnseq_v1b_fitness_vs_moi.pdf", p1, width = 10, height = 5)


# ── Plot 2: Essential genes — complementation rescue vs MOI ──────────────────
p2 <- results_all %>%
  filter(gene_class == "Essential") %>%
  ggplot(aes(x = factor(moi), y = fitness_score,
             color = complementation, group = gene_id)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  scale_color_viridis_c(name = "Complementation\nFraction", option = "plasma") +
  #geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  labs(
    title    = "Essential Genes: Apparent Fitness vs MOI",
    # subtitle = "complementable genes increase in abundance with high MOI",
    x        = "MOI",
    y        = "Normalized log2 abundance"
  ) +
  theme_bw(base_size = 12)

print(p2)
#ggsave("tnseq_v1b_essential_genes.pdf", p2, width = 8, height = 5)


# ── Plot 3: Analtical helper probability overlaid on simulation ──────────────
# Average fitness score of essential genes vs MOI alongside P(helper)
# I dont like thi plot.  

# essential_summary <- results_all %>%
#   filter(gene_class == "Essential") %>%
#   group_by(moi) %>%
#   summarise(mean_fitness_score = mean(fitness_score), .groups = "drop") %>%
#   mutate(p_helper = p_at_least_one_helper(moi, n_genes = 40))
# 
# # #p3 <- ggplot(essential_summary, aes(x = moi)) +
# #   geom_line(aes(y = mean_fitness_score, color = "Mean fitness score (essential genes)"),
# #             linewidth = 1.2) +
# #   geom_point(aes(y = mean_fitness_score, color = "Mean fitness score (essential genes)"),
# #              size = 3) +
# #   geom_line(aes(y = p_helper * max(abs(mean_fitness_score)) * -1,
# #                 color = "P(at least one helper) [scaled]"),
# #             linewidth = 1.2, linetype = "dashed") +
# #   geom_point(aes(y = p_helper * max(abs(mean_fitness_score)) * -1,
# #                  color = "P(at least one helper) [scaled]"), size = 3) +
# #   scale_color_manual(values = c(
# #     "Mean fitness score (essential genes)" = "#d62728",
# #     "P(at least one helper) [scaled]"      = "#377eb8"
# #   )) +
# #   labs(
# #     title    = "Complementation Probability Drives MOI Artifact",
# #     # subtitle = "As MOI rises, P(helper present) increases, inflating essential gene scores",
# #     x        = "MOI",
# #     y        = "Value (fitness score)",
# #     color    = NULL
# #   ) +
# #   theme_bw(base_size = 12) +
# #   theme(legend.position = "bottom")
# # 
# # #print(p3)
# # ggsave("tnseq_v1b_helper_probability.pdf", p3, width = 8, height = 5)
# # 
# 
# # ── Plot 4: True vs estimated fitness at three MOIs ───────────────────────────
# p4 <- results_all %>%
#   filter(moi %in% c(0.1, 1, 5)) %>%
#   mutate(moi_label = paste0("MOI = ", moi)) %>%
#   ggplot(aes(x = true_fitness, y = fitness_score,
#              color = complementation, shape = gene_class)) +
#   geom_point(size = 2.5, alpha = 0.85) +
#   geom_smooth(method = "lm", se = FALSE, color = "black", linewidth = 0.8) +
#   facet_wrap(~moi_label) +
#   scale_color_viridis_c(name = "Complementation", option = "plasma") +
#   scale_shape_manual(values = c(Essential = 17, Moderate = 16, "Near-neutral" = 1)) +
#   labs(
#     title    = "True vs Estimated Fitness at Different MOIs",
#     subtitle = "Score compression at high MOI is strongest for high-complementation genes",
#     x        = "True fitness (simulation parameter)",
#     y        = "Estimated fitness score",
#     shape    = "Gene class"
#   ) +
#   theme_bw(base_size = 12)
# 
# print(p4)
# ggsave("tnseq_v1b_true_vs_estimated.pdf", p4, width = 11, height = 4)


# ─────────────────────────────────────────────────────────────────────────────
# 8. SUMMARY STATISTICS
# ─────────────────────────────────────────────────────────────────────────────
# 
# dont really need these but they might be usful diagnostics. 

# cat("\n══════════════════════════════════════════════════════\n")
# cat("Correlation: true fitness vs estimated fitness score\n")
# cat("══════════════════════════════════════════════════════\n")
# 
# results_all %>%
#   group_by(moi) %>%
#   summarise(
#     pearson_r    = round(cor(true_fitness, fitness_score, method = "pearson"),  3),
#     spearman_rho = round(cor(true_fitness, fitness_score, method = "spearman"), 3),
#     .groups = "drop"
#   ) %>%
#   print()
# 
# cat("\n══════════════════════════════════════════════════════\n")
# cat("Mean fitness score for essential genes by MOI\n")
# cat("══════════════════════════════════════════════════════\n")
# 
# results_all %>%
#   filter(gene_class == "Essential") %>%
#   group_by(moi) %>%
#   summarise(
#     mean_score = round(mean(fitness_score), 3),
#     sd_score   = round(sd(fitness_score),   3),
#     .groups    = "drop"
#   ) %>%
#   print()
```

Figure 4 compartment split model with random draw. 

```{r}
################################################################################
# TnSeq Phage Simulation - : Compartmentalized Life Cycle + Saturated Mutant Library
#
#
# Life cycle stages (33 genes each):
#   1. ADSORPTION   — gene disruption = phage never enters the cell.
#                     Complementation always 0 (cis requirement, pre-entry).
#
#   2. REPLICATION  — intracellular; products are diffusible.
#                     Co-infectors can complement if WT at this gene.
#
#   3. EXIT (lysis) — assembly and release.
#                     Partial complementation only (cap at 0.7): some
#                     structural genes must be cis-encoded for packaging.
################################################################################

library(tidyverse)

# ─────────────────────────────────────────────────────────────────────────────
# 1. GENE PARAMETER TABLE
# ─────────────────────────────────────────────────────────────────────────────

set.seed(42)

n_genes <- 99

# 33 genes per stage
stage_assignment <- c(
  rep("adsorption",  33),
  rep("replication", 33),
  rep("exit",        33)
)

gene_params <- tibble(
  gene_id = paste0("gene_", sprintf("%03d", 1:n_genes)),
  stage   = sample(stage_assignment),   # shuffle so stage isn't confounded with gene order

  # True fitness effect of disruption (0 = lethal, 1 = fully neutral)
  # Equal thirds: 33 essential, 33 moderate, 33 near-neutral
  fitness_effect = c(
    runif(33, 0.00, 0.10),   # 33 essential
    runif(33, 0.10, 0.60),   # 33 moderate
    runif(33, 0.70, 1.00)    # 33 near-neutral
  ) %>% sample(),

  # Raw complementation drawn in equal thirds before stage constraint applied
  complementation_raw = c(
    runif(33, 0.00, 0.10),   # 33 low complementation
    runif(33, 0.40, 0.80),   # 33 moderate complementation
    runif(33, 0.85, 1.00)    # 33 high complementation
  ) %>% sample()

) %>%
  mutate(
    # Stage constrains realised complementation:
    #   adsorption  → always 0 (phage never entered, no trans-complementation possible)
    #   replication → full range allowed
    #   exit        → capped at 0.7 (some structural genes need cis encoding)
    complementation = case_when(
      stage == "adsorption"  ~ 0,
      stage == "replication" ~ complementation_raw,
      stage == "exit"        ~ complementation_raw * 0.7
    )
  ) %>%
  select(-complementation_raw)

cat("Gene stage distribution:\n")
print(table(gene_params$stage))

cat("\nEssentiality distribution:\n")
print(table(cut(gene_params$fitness_effect,
               breaks = c(0, 0.1, 0.6, 1),
               labels = c("Essential", "Moderate", "Near-neutral"),
               include.lowest = TRUE)))

cat("\nGene parameters (first 15 rows):\n")
print(gene_params, n = 15)


# ─────────────────────────────────────────────────────────────────────────────
# 2. STAGE-AWARE LIFE CYCLE FITNESS FUNCTION
# ─────────────────────────────────────────────────────────────────────────────

#' Calculate relative phage output for one mutant phage in one cell.
#'
#' Helper counting is stage-aware:
#'   - Adsorption gene mutant: cannot be complemented regardless of co-infectors.
#'   - Replication / exit gene mutant: helpers are co-infectors whose insertion
#'     is in a DIFFERENT gene (they carry a functional copy of the focal gene).
#'
#' @param focal_gene       Integer row index of the disrupted gene
#' @param n_helpers        Number of co-infectors WT at the focal gene
#'                         (pre-computed outside this function)
#' @param gene_params      Gene parameter tibble
#' @return Relative phage output (0–1)

phage_lifecycle <- function(focal_gene, n_helpers, gene_params) {

  gp          <- gene_params[focal_gene, ]
  intrinsic   <- gp$fitness_effect
  has_helper  <- n_helpers > 0

  if (gp$stage == "adsorption") {
    # Phage cannot enter cell — no complementation possible regardless of MOI
    return(intrinsic)
  }

  if (gp$stage == "replication") {
    if (has_helper) {
      return(intrinsic + gp$complementation * (1 - intrinsic))
    } else {
      return(intrinsic)
    }
  }

  if (gp$stage == "exit") {
    # Assume replication succeeded (exit gene disruption doesn't block replication)
    if (has_helper) {
      return(intrinsic + gp$complementation * (1 - intrinsic))
    } else {
      return(intrinsic)
    }
  }

  return(1.0)  # fallback
}


# ─────────────────────────────────────────────────────────────────────────────
# 3. ANALYTICAL HELPER PROBABILITY
# ─────────────────────────────────────────────────────────────────────────────

#' P(at least one co-infector is WT at the focal gene)
#' given Poisson MOI and uniform insertion distribution across n_genes.
#'
#' Each co-infector has probability (1 - 1/n_genes) of being WT at focal gene.
#' Result from Poisson probability generating function:
#'   P(at least one helper) = 1 - exp(-MOI * (1 - 1/n_genes))
#'
#' Note: adsorption genes are excluded — this probability is only meaningful
#' for replication and exit genes.

p_at_least_one_helper <- function(moi, n_genes) {
  1 - exp(-moi * (1 - 1 / n_genes))
}

cat("\nP(at least one helper) for replication/exit genes | n_genes = 99:\n")
tibble(moi = c(0.1, 0.5, 1, 2, 5, 10)) %>%
  mutate(p_helper = round(p_at_least_one_helper(moi, n_genes = 99), 3)) %>%
  print()


# ─────────────────────────────────────────────────────────────────────────────
# 4. CORE SIMULATION — SATURATED MUTANT LIBRARY
# ─────────────────────────────────────────────────────────────────────────────

#' Simulate one TnSeq experiment
#'
#' Every phage carries an insertion in one randomly chosen gene (saturated library).
#' For each focal mutant phage in a cell, helpers are co-infecting phages whose
#' insertion is in a DIFFERENT gene — they supply a functional copy of the focal gene.
#' Adsorption mutants receive no benefit from co-infectors.
#'
#' @param moi          Mean MOI (Poisson lambda)
#' @param n_bacteria   Number of bacteria
#' @param burst_size   Mean WT burst size
#' @param gene_params  Gene parameter tibble
#' @param seed         Random seed
#' @return Tibble with per-gene TnSeq fitness estimates

simulate_tnseq_v2b <- function(moi         = 1,
                                n_bacteria  = 2e5,
                                burst_size  = 50,
                                gene_params = gene_params,
                                seed        = NULL) {

  if (!is.null(seed)) set.seed(seed)
  n_genes <- nrow(gene_params)

  input_counts  <- numeric(n_genes)
  output_counts <- numeric(n_genes)

  # Draw phages per cell
  phages_per_cell <- rpois(n_bacteria, lambda = moi)
  phages_per_cell <- phages_per_cell[phages_per_cell > 0]
  n_infected      <- length(phages_per_cell)

  for (cell in seq_len(n_infected)) {
    n_phage <- phages_per_cell[cell]

    # Every phage has an insertion — assign each to a random gene
    cell_genes <- sample(n_genes, n_phage, replace = TRUE)

    for (p in seq_len(n_phage)) {
      focal_gene       <- cell_genes[p]
      coinfector_genes <- cell_genes[-p]

      input_counts[focal_gene] <- input_counts[focal_gene] + 1

      # Count co-infectors that are WT at the focal gene
      # (adsorption gene: n_helpers is computed but ignored inside phage_lifecycle)
      n_helpers <- sum(coinfector_genes != focal_gene)

      rel_fitness   <- phage_lifecycle(focal_gene, n_helpers, gene_params)
      actual_output <- rpois(1, lambda = max(rel_fitness * burst_size, 1e-9))

      output_counts[focal_gene] <- output_counts[focal_gene] + actual_output
    }
  }

  # Estimate fitness score
  pseudo          <- 1
  log2_ratio      <- log2((output_counts + pseudo) / (input_counts + pseudo))
  neutral_idx     <- which(gene_params$fitness_effect > 0.7)
  log2_ratio_norm <- log2_ratio - median(log2_ratio[neutral_idx])

  tibble(
    gene_id         = gene_params$gene_id,
    stage           = gene_params$stage,
    true_fitness    = gene_params$fitness_effect,
    complementation = gene_params$complementation,
    input_counts    = input_counts,
    output_counts   = output_counts,
    log2_ratio      = log2_ratio,
    fitness_score   = log2_ratio_norm,
    moi             = moi
  )
}


# ─────────────────────────────────────────────────────────────────────────────
# 5. RUN ACROSS MOIs
# ─────────────────────────────────────────────────────────────────────────────

moi_values <- c(0.1, 0.5, 1, 2, 5, 10)

cat("\nRunning simulations...\n")
results_all <- map_dfr(moi_values, function(m) {
  cat("  MOI =", m, "\n")
  simulate_tnseq_v2b(
    moi         = m,
    n_bacteria  = 2e5,
    burst_size  = 50,
    gene_params = gene_params,
    seed        = 99
  )
})


# ─────────────────────────────────────────────────────────────────────────────
# 6. VISUALIZATIONS
# ─────────────────────────────────────────────────────────────────────────────

results_all <- results_all %>%
  mutate(
    essentiality = case_when(
      true_fitness < 0.1  ~ "Essential",
      true_fitness < 0.6  ~ "Moderate",
      TRUE                ~ "Near-neutral"
    ),
    stage = factor(stage, levels = c("adsorption", "replication", "exit"))
  )

stage_colors <- c(adsorption  = "#e41a1c",
                  replication = "#377eb8",
                  exit        = "#4daf4a")

# ── Plot 1: Fitness score vs MOI, faceted by stage × essentiality ────────────
p1 <- results_all %>%
  #filter(essentiality != "Near-neutral") %>%
  ggplot(aes(x = factor(moi), y = fitness_score,
             color = stage, group = gene_id)) +
  geom_line(alpha = 0.45, linewidth = 0.5) +
  geom_point(alpha = 0.65, size = 1.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  facet_grid(essentiality ~ stage) +
  scale_color_manual(values = stage_colors) +
  labs(
    title    = "TnSeq Fitness Score vs MOI — Compartmentalized Life Cycle (Saturated Library)",
    subtitle = "Essential and moderate genes; each line = one gene tracked across MOIs",
    x        = "MOI",
    y        = "Normalized log2(output / input)",
    color    = "Life cycle stage"
  ) +
  theme_bw(base_size = 11) +
  theme(legend.position = "bottom", strip.text = element_text(face = "bold"))

print(p1)
ggsave("tnseq_v2b_fitness_by_stage_moi.pdf", p1, width = 11, height = 7)


# ── Plot 2: Essential genes only — color by complementation ──────────────────
p2 <- results_all %>%
  filter(essentiality == "Essential") %>%
  ggplot(aes(x = factor(moi), y = fitness_score,
             color = complementation, group = gene_id)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2.2) +
  facet_wrap(~stage, nrow = 1) +
  scale_color_viridis_c(name = "Complementation\nFraction", option = "plasma") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  labs(
    title    = "Essential Genes: Apparent Fitness vs MOI by Stage",
    subtitle = "Adsorption genes flat regardless of MOI; replication/exit genes rescued if complementable",
    x        = "MOI",
    y        = "Normalized log2(output / input)"
  ) +
  theme_bw(base_size = 11)

print(p2)
ggsave("tnseq_v2b_essential_by_stage.pdf", p2, width = 11, height = 5)


# ── Plot 3: True vs estimated fitness at MOI 0.1, 1, 5 ───────────────────────
p3 <- results_all %>%
  filter(moi %in% c(0.1, 1, 5)) %>%
  mutate(moi_label = paste0("MOI = ", moi)) %>%
  ggplot(aes(x = true_fitness, y = fitness_score,
             color = stage, shape = essentiality)) +
  geom_point(size = 2.2, alpha = 0.8) +
  geom_smooth(method = "lm", se = FALSE, color = "black",
              linewidth = 0.7, aes(group = moi_label)) +
  facet_wrap(~moi_label, nrow = 1) +
  scale_color_manual(values = stage_colors) +
  scale_shape_manual(values = c(Essential = 17, Moderate = 16, "Near-neutral" = 1)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey60") +
  labs(
    title    = "True vs Estimated Fitness — Saturated Library",
    subtitle = "Score compression at high MOI; adsorption genes (red) unaffected",
    x        = "True fitness (simulation parameter)",
    y        = "Estimated fitness score (normalized log2 ratio)",
    color    = "Stage", shape = "Essentiality"
  ) +
  theme_bw(base_size = 11) +
  theme(legend.position = "right")

print(p3)
ggsave("tnseq_v2b_true_vs_estimated.pdf", p3, width = 13, height = 5)


# ── Plot 4: Heatmap — all 99 genes × all MOIs ────────────────────────────────
heat_data <- results_all %>%
  arrange(stage, true_fitness) %>%
  mutate(gene_id = factor(gene_id, levels = unique(gene_id)))

p4 <- ggplot(heat_data, aes(x = factor(moi), y = gene_id, fill = fitness_score)) +
  geom_tile(color = "white", linewidth = 0.2) +
  scale_fill_gradient2(low = "#d73027", mid = "white", high = "#1a9641",
                       midpoint = 0, name = "Fitness\nScore") +
  facet_grid(stage ~ ., scales = "free_y", space = "free") +
  labs(
    title    = "TnSeq Fitness Scores: All 99 Genes × All MOIs",
    subtitle = "Sorted within stage by true fitness (most essential = top of each panel)",
    x        = "MOI",
    y        = "Gene"
  ) +
  theme_bw(base_size = 9) +
  theme(
    axis.text.y      = element_text(size = 5),
    strip.text.y     = element_text(angle = 0, face = "bold"),
    legend.position  = "right"
  )

print(p4)
ggsave("tnseq_v2b_heatmap.pdf", p4, width = 9, height = 14)


# ── Plot 5: Analytical helper probability by MOI — adsorption excluded ────────
helper_curve <- tibble(moi = seq(0.1, 10, by = 0.1)) %>%
  mutate(p_helper = p_at_least_one_helper(moi, n_genes = 99))

# Overlay mean fitness score of essential replication + exit genes
essential_rep_exit <- results_all %>%
  filter(essentiality == "Essential", stage != "adsorption") %>%
  group_by(moi) %>%
  summarise(mean_score = mean(fitness_score), .groups = "drop")

p5 <- ggplot() +
  geom_line(data = helper_curve,
            aes(x = moi, y = p_helper, linetype = "P(helper present)"),
            color = "#377eb8", linewidth = 1.2) +
  geom_line(data = essential_rep_exit,
            aes(x = moi, y = (mean_score - min(mean_score)) /
                  (max(mean_score) - min(mean_score)),
                linetype = "Essential rep/exit score\n(scaled 0–1)"),
            color = "#e41a1c", linewidth = 1.2) +
  geom_point(data = essential_rep_exit,
             aes(x = moi, y = (mean_score - min(mean_score)) /
                   (max(mean_score) - min(mean_score))),
             color = "#e41a1c", size = 3) +
  scale_linetype_manual(values = c("P(helper present)" = "solid",
                                   "Essential rep/exit score\n(scaled 0–1)" = "dashed"),
                        name = NULL) +
  labs(
    title    = "Complementation Probability Drives the MOI Artifact",
    subtitle = "P(helper present) rises with MOI, inflating apparent fitness of essential replication/exit genes",
    x        = "MOI",
    y        = "Scaled value (0–1)"
  ) +
  theme_bw(base_size = 12) +
  theme(legend.position = "bottom")

print(p5)
ggsave("tnseq_v2b_helper_probability.pdf", p5, width = 8, height = 5)


# ─────────────────────────────────────────────────────────────────────────────
# 7. SUMMARY STATISTICS
# ─────────────────────────────────────────────────────────────────────────────

cat("\n══════════════════════════════════════════════════════════════\n")
cat("Correlation: true fitness vs estimated fitness score by stage\n")
cat("══════════════════════════════════════════════════════════════\n")

results_all %>%
  group_by(moi, stage) %>%
  summarise(
    pearson_r    = round(cor(true_fitness, fitness_score, method = "pearson"),  3),
    spearman_rho = round(cor(true_fitness, fitness_score, method = "spearman"), 3),
    n_genes      = n(),
    .groups      = "drop"
  ) %>%
  print(n = 30)

cat("\n══════════════════════════════════════════════════════════════\n")
cat("Mean fitness score for ESSENTIAL genes by stage and MOI\n")
cat("══════════════════════════════════════════════════════════════\n")

results_all %>%
  filter(essentiality == "Essential") %>%
  group_by(moi, stage) %>%
  summarise(mean_score = round(mean(fitness_score), 3), .groups = "drop") %>%
  pivot_wider(names_from = stage, values_from = mean_score) %>%
  print()
```

figure 5
```{r}
################################################################################
# TnSeq Phage Simulation: Two-Round Infection
#
# Experimental design:
#   ROUND 1 — Standard TnSeq infection at a defined MOI.
#             Co-infection is common; complementation can mask fitness defects.
#             Produces an output phage pool with altered gene frequencies.
#
#   ROUND 2 — Each phage from the Round 1 output pool infects a new host ALONE
#             (effective MOI → 0; single phage per cell, no co-infectors).
#             No complementation possible. Fitness is purely intrinsic.
#             Produces a second output pool.
#
# This design separates:
#   - GLOBAL competition  : overall frequency change across both rounds
#   - LOCAL competition   : within-cell complementation in Round 1 only
#   - COMPLEMENTATION MASK: genes that appeared neutral in Round 1 because they
#                           were rescued, but crash in Round 2 when isolated
#
# The key metric is the DELTA score:
#   delta = fitness_score_round2 - fitness_score_round1
#   Negative delta → gene was being complemented/rescued in Round 1
#   Near-zero delta → gene fitness is MOI-independent (adsorption genes,
#                     or genuinely neutral genes)
#
# Population structure notes:
#   - Round 1 output pool frequencies are the Round 2 INPUT frequencies.
#     Mutants that were rescued in Round 1 enter Round 2 at HIGHER frequency
#     than their intrinsic fitness warrants — this amplifies the Round 2 crash.
#   - Adsorption mutants: fail at same rate in both rounds (no MOI effect),
#     so delta ≈ 0. Distinguishable from complemented replication/exit mutants.
################################################################################

library(tidyverse)

# ─────────────────────────────────────────────────────────────────────────────
# 1. GENE PARAMETER TABLE  (identical to v2b)
# ─────────────────────────────────────────────────────────────────────────────

set.seed(42)

n_genes <- 99

stage_assignment <- c(
  rep("adsorption",  33),
  rep("replication", 33),
  rep("exit",        33)
)

gene_params <- tibble(
  gene_id = paste0("gene_", sprintf("%03d", 1:n_genes)),
  stage   = sample(stage_assignment),

  fitness_effect = c(
    runif(33, 0.00, 0.10),
    runif(33, 0.10, 0.60),
    runif(33, 0.70, 1.00)
  ) %>% sample(),

  complementation_raw = c(
    runif(33, 0.00, 0.10),
    runif(33, 0.40, 0.80),
    runif(33, 0.85, 1.00)
  ) %>% sample()

) %>%
  mutate(
    complementation = case_when(
      stage == "adsorption"  ~ 0,
      stage == "replication" ~ complementation_raw,
      stage == "exit"        ~ complementation_raw
    )
  ) %>%
  select(-complementation_raw)

cat("Gene stage distribution:\n")
print(table(gene_params$stage))


# ─────────────────────────────────────────────────────────────────────────────
# 2. LIFE CYCLE FITNESS FUNCTION  (unchanged from v2b)
# ─────────────────────────────────────────────────────────────────────────────

phage_lifecycle <- function(focal_gene, n_helpers, gene_params) {
  gp         <- gene_params[focal_gene, ]
  intrinsic  <- gp$fitness_effect
  has_helper <- n_helpers > 0

  # -- Adsorption: never complemented -----------------------------------------
  if (gp$stage == "adsorption") return(intrinsic)

  # -- Replication: complementable by co-infectors ----------------------------
  if (gp$stage == "replication") {
    if (has_helper) {
      return(intrinsic + gp$complementation * (1 - intrinsic))
    } else {
      return(intrinsic)
    }
  }

  # -- Exit: phage alone cannot release progeny -- burst goes to zero ----------
  # Biology: even if the genome replicates normally, without functional exit
  # machinery the cell never lyses and no progeny are released.
  # A co-infector can supply exit products in trans (up to complementation limit).
  if (gp$stage == "exit") {
    if (!has_helper) return(0)
    return(intrinsic + gp$complementation * (1 - intrinsic))
  }

  return(1.0)  # fallback
}


# ─────────────────────────────────────────────────────────────────────────────
# 3. SINGLE-ROUND SIMULATION ENGINE
# ─────────────────────────────────────────────────────────────────────────────

#' Run one round of infection.
#'
#' Input is either a flat saturated library (Round 1) or a frequency-weighted
#' pool from the previous round's output (Round 2+).
#'
#' @param moi              Mean MOI for this round (Poisson lambda)
#' @param n_bacteria       Number of bacteria
#' @param burst_size       Mean WT burst size
#' @param gene_params      Gene parameter tibble
#' @param input_freq       Named numeric vector of relative input frequencies
#'                         per gene (length n_genes, sums to 1).
#'                         If NULL, assumes uniform (saturated library).
#' @param seed             Random seed
#' @return List with:
#'   $results   tibble of per-gene counts and scores
#'   $output_freq  named vector of output frequencies (for chaining rounds)

run_infection_round <- function(moi         = 1,
                                n_bacteria  = 2e5,
                                burst_size  = 50,
                                gene_params = gene_params,
                                input_freq  = NULL,
                                seed        = NULL) {

  if (!is.null(seed)) set.seed(seed)
  n_genes <- nrow(gene_params)

  # Default: uniform input frequencies (saturated library)
  if (is.null(input_freq)) {
    input_freq <- rep(1 / n_genes, n_genes)
  }
  input_freq <- input_freq / sum(input_freq)  # ensure normalised

  input_counts  <- numeric(n_genes)
  output_counts <- numeric(n_genes)

  phages_per_cell <- rpois(n_bacteria, lambda = moi)
  phages_per_cell <- phages_per_cell[phages_per_cell > 0]
  n_infected      <- length(phages_per_cell)

  for (cell in seq_len(n_infected)) {
    n_phage    <- phages_per_cell[cell]
    cell_genes <- sample(n_genes, n_phage, replace = TRUE, prob = input_freq)

    for (p in seq_len(n_phage)) {
      focal_gene       <- cell_genes[p]
      coinfector_genes <- cell_genes[-p]

      input_counts[focal_gene] <- input_counts[focal_gene] + 1

      n_helpers     <- sum(coinfector_genes != focal_gene)
      rel_fitness   <- phage_lifecycle(focal_gene, n_helpers, gene_params)
      actual_output <- rpois(1, lambda = max(rel_fitness * burst_size, 1e-9))

      output_counts[focal_gene] <- output_counts[focal_gene] + actual_output
    }
  }

  # Fitness score
  pseudo          <- 1
  log2_ratio      <- log2((output_counts + pseudo) / (input_counts + pseudo))
  neutral_idx     <- which(gene_params$fitness_effect > 0.7)
  log2_ratio_norm <- log2_ratio - median(log2_ratio[neutral_idx])

  # Output frequencies for chaining into next round
  output_freq <- (output_counts + pseudo) / sum(output_counts + pseudo)

  results <- tibble(
    gene_id         = gene_params$gene_id,
    stage           = gene_params$stage,
    true_fitness    = gene_params$fitness_effect,
    complementation = gene_params$complementation,
    input_counts    = input_counts,
    output_counts   = output_counts,
    log2_ratio      = log2_ratio,
    fitness_score   = log2_ratio_norm
  )

  list(results = results, output_freq = output_freq)
}


# ─────────────────────────────────────────────────────────────────────────────
# 4. TWO-ROUND EXPERIMENT
# ─────────────────────────────────────────────────────────────────────────────

#' Run the full two-round experiment for one Round 1 MOI value.
#'
#' Round 1: standard co-infection at moi_r1
#' Round 2: each phage infects alone — effective MOI forced to ~0
#'          (implemented as moi = 0.01 so cells receive at most 1 phage
#'           with very high probability; Poisson(0.01) gives P(≥2) < 0.005%)
#'
#' @param moi_r1      MOI for Round 1
#' @param moi_r2      MOI for Round 2 (default 0.01 ≈ single infection)
#' @param n_bacteria  Bacteria per round
#' @param burst_size  WT burst size
#' @param gene_params Gene parameter tibble
#' @param seed        Base random seed
#' @return Tibble with per-gene scores for both rounds and delta

two_round_experiment <- function(moi_r1      = 1,
                                 moi_r2      = 0.01,
                                 n_bacteria  = 2e5,
                                 burst_size  = 50,
                                 gene_params = gene_params,
                                 seed        = 99) {

  # Round 1 — co-infection, uniform input library
  r1 <- run_infection_round(
    moi        = moi_r1,
    n_bacteria = n_bacteria,
    burst_size = burst_size,
    gene_params= gene_params,
    input_freq = NULL,
    seed       = seed
  )

  # Round 2 — single infection, input frequencies from Round 1 output
  r2 <- run_infection_round(
    moi        = moi_r2,
    n_bacteria = n_bacteria,
    burst_size = burst_size,
    gene_params= gene_params,
    input_freq = r1$output_freq,
    seed       = seed + 1
  )

  # Combine and compute delta
  r1$results %>%
    select(gene_id, stage, true_fitness, complementation,
           fitness_score_r1 = fitness_score) %>%
    left_join(
      r2$results %>% select(gene_id, fitness_score_r2 = fitness_score),
      by = "gene_id"
    ) %>%
    mutate(
      delta       = fitness_score_r2 - fitness_score_r1,
      moi_r1      = moi_r1,
      moi_r2      = moi_r2
    )
}


# ─────────────────────────────────────────────────────────────────────────────
# 5. RUN ACROSS MULTIPLE ROUND 1 MOIs
# ─────────────────────────────────────────────────────────────────────────────

moi_r1_values <- c(0.1, 0.5, 1, 2, 5, 10)

cat("\nRunning two-round experiments across Round 1 MOIs...\n")

results_all <- map_dfr(moi_r1_values, function(m) {
  cat("  Round 1 MOI =", m, "\n")
  two_round_experiment(
    moi_r1     = m,
    moi_r2     = 0.01,
    n_bacteria = 2e5,
    burst_size = 50,
    gene_params= gene_params,
    seed       = 99
  )
})

# Add classification labels
results_all <- results_all %>%
  mutate(
    essentiality = case_when(
      true_fitness < 0.1  ~ "Essential",
      true_fitness < 0.6  ~ "Moderate",
      TRUE                ~ "Near-neutral"
    ),
    stage = factor(stage, levels = c("adsorption", "replication", "exit"))
  )

stage_colors <- c(adsorption  = "#e41a1c",
                  replication = "#377eb8",
                  exit        = "#4daf4a")


# ─────────────────────────────────────────────────────────────────────────────
# 6. VISUALIZATIONS
# ─────────────────────────────────────────────────────────────────────────────

# ── Plot 1: Round 1 vs Round 2 fitness scores, essential genes only ───────────
# Shows the "unmasking" of complemented mutants in Round 2
p1 <- results_all %>%
  filter(essentiality == "Essential") %>%
  pivot_longer(cols = c(fitness_score_r1, fitness_score_r2),
               names_to = "round",
               values_to = "fitness_score") %>%
  mutate(round = recode(round,
                        fitness_score_r1 = "Round 1 (co-infection)",
                        fitness_score_r2 = "Round 2 (single infection)")) %>%
  ggplot(aes(x = round, y = fitness_score,
             color = stage, group = gene_id)) +
  geom_line(alpha = 0.5, linewidth = 0.6) +
  geom_point(alpha = 0.8, size = 1.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  facet_grid(stage ~ factor(moi_r1, labels = paste0("R1 MOI=", moi_r1_values))) +
  scale_color_manual(values = stage_colors) +
  labs(
    title    = "Essential Gene Fitness: Round 1 (co-infection) vs Round 2 (single infection)",
    subtitle = "Replication/exit mutants rescued in Round 1 crash in Round 2; adsorption mutants unchanged",
    x        = NULL,
    y        = "Normalized fitness score",
    color    = "Stage"
  ) +
  theme_bw(base_size = 10) +
  theme(axis.text.x  = element_text(angle = 25, hjust = 1),
        legend.position = "bottom",
        strip.text = element_text(face = "bold", size = 8))

print(p1)
ggsave("tnseq_v3_round1_vs_round2_essential.pdf", p1, width = 13, height = 9)


# ── Plot 2: Delta score (R2 - R1) by stage and complementation ───────────────
# The delta is the complementation mask: negative = was being rescued in R1
p2 <- results_all %>%
  # filter(essentiality != "Near-neutral") %>%
  ggplot(aes(x = factor(moi_r1), y = delta,
             color = complementation, group = gene_id)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_line(alpha = 0.4, linewidth = 0.5) +
  geom_point(alpha = 0.8, size = 1.8) +
  facet_grid(essentiality ~ stage) +
  scale_color_viridis_c(name = "Complementation\nFraction", option = "plasma") +
  labs(
    title    = "Delta Score (Round 2 − Round 1) by Stage and Essentiality",
    subtitle = "Negative delta = gene was complemented/rescued in Round 1; adsorption genes cluster near zero",
    x        = "Round 1 MOI",
    y        = "Delta fitness score (R2 − R1)"
  ) +
  theme_bw(base_size = 11) +
  theme(strip.text = element_text(face = "bold"))

print(p2)
ggsave("tnseq_v3_delta_by_stage.pdf", p2, width = 12, height = 7)


# ── Plot 3: Delta vs complementation fraction — the core diagnostic plot ──────
# At high R1 MOI, delta should correlate strongly with complementation
# for replication/exit genes; adsorption genes should be flat near zero
p3 <- results_all %>%
  filter(essentiality == "Essential",
         moi_r1 %in% c(0.1, 1, 5, 10)) %>%
  mutate(moi_label = paste0("R1 MOI = ", moi_r1)) %>%
  ggplot(aes(x = complementation, y = delta, color = stage, shape = stage)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_point(size = 3, alpha = 0.85) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 0.8,
              aes(group = stage), alpha = 0.15) +
  facet_wrap(~moi_label, nrow = 1) +
  scale_color_manual(values = stage_colors) +
  scale_shape_manual(values = c(adsorption = 17, replication = 16, exit = 15)) +
  labs(
    title    = "Delta Score vs Complementation Fraction — Essential Genes",
    subtitle = "Strong negative correlation for replication/exit at high MOI; adsorption genes flat",
    x        = "Complementation fraction",
    y        = "Delta fitness score (R2 − R1)",
    color    = "Stage", shape = "Stage"
  ) +
  theme_bw(base_size = 11) +
  theme(legend.position = "bottom")

print(p3)
ggsave("tnseq_v3_delta_vs_complementation.pdf", p3, width = 13, height = 5)


# ── Plot 4: Population structure — input vs output frequencies across rounds ──
# Shows how the phage population shifts: complemented mutants inflate in R1,
# then are purged in R2
freq_data <- map_dfr(moi_r1_values, function(m) {

  r1 <- run_infection_round(
    moi = m, n_bacteria = 2e5, burst_size = 50,
    gene_params = gene_params, input_freq = NULL, seed = 99
  )
  r2 <- run_infection_round(
    moi = 0.01, n_bacteria = 2e5, burst_size = 50,
    gene_params = gene_params, input_freq = r1$output_freq, seed = 100
  )

  tibble(
    gene_id      = gene_params$gene_id,
    stage        = gene_params$stage,
    true_fitness = gene_params$fitness_effect,
    complementation = gene_params$complementation,
    freq_input   = rep(1 / n_genes, n_genes),
    freq_r1_out  = r1$output_freq,
    freq_r2_out  = r2$output_freq,
    moi_r1       = m
  )
}) %>%
  mutate(
    enrichment_r1 = log2(freq_r1_out / freq_input),
    enrichment_r2 = log2(freq_r2_out / freq_r1_out),
    essentiality  = case_when(
      true_fitness < 0.1  ~ "Essential",
      true_fitness < 0.6  ~ "Moderate",
      TRUE                ~ "Near-neutral"
    ),
    stage = factor(stage, levels = c("adsorption", "replication", "exit"))
  )

p4 <- freq_data %>%
  filter(essentiality == "Essential", moi_r1 %in% c(1, 5, 10)) %>%
  ggplot(aes(x = enrichment_r1, y = enrichment_r2,
             color = complementation, shape = stage)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey60") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey60") +
  geom_point(size = 3, alpha = 0.9) +
  facet_wrap(~paste0("R1 MOI = ", moi_r1), nrow = 1) +
  scale_color_viridis_c(name = "Complementation", option = "plasma") +
  scale_shape_manual(values = c(adsorption = 17, replication = 16, exit = 15)) +
  labs(
    title    = "Population Structure: Round 1 Enrichment vs Round 2 Purging",
    subtitle = "Top-left = rescued in R1, purged in R2 (high complementation replication/exit genes)\nBottom-left = essential regardless (adsorption, or low complementation)",
    x        = "log2 enrichment in Round 1 output vs input",
    y        = "log2 enrichment in Round 2 output vs R1 output",
    shape    = "Stage"
  ) +
  theme_bw(base_size = 11) +
  theme(legend.position = "right")

print(p4)
ggsave("tnseq_v3_population_structure.pdf", p4, width = 13, height = 5)


# ── Plot 5: Heatmap — delta scores for all genes across all R1 MOIs ───────────
heat_data <- results_all %>%
  arrange(stage, true_fitness) %>%
  mutate(gene_id = factor(gene_id, levels = unique(gene_id)))

p5 <- ggplot(heat_data, aes(x = factor(moi_r1), y = gene_id, fill = delta)) +
  geom_tile(color = "white", linewidth = 0.2) +
  scale_fill_gradient2(low = "#d73027", mid = "white", high = "#1a9641",
                       midpoint = 0, name = "Delta\n(R2−R1)") +
  facet_grid(stage ~ ., scales = "free_y", space = "free") +
  labs(
    title    = "Delta Fitness Score (R2 − R1) Heatmap",
    subtitle = "Red = gene was rescued in R1, exposed in R2 | Sorted by true fitness within stage",
    x        = "Round 1 MOI",
    y        = "Gene"
  ) +
  theme_bw(base_size = 9) +
  theme(
    axis.text.y      = element_text(size = 5),
    strip.text.y     = element_text(angle = 0, face = "bold"),
    legend.position  = "right"
  )

print(p5)
ggsave("tnseq_v3_delta_heatmap.pdf", p5, width = 9, height = 14)


# ─────────────────────────────────────────────────────────────────────────────
# 7. SUMMARY STATISTICS
# ─────────────────────────────────────────────────────────────────────────────

cat("\n══════════════════════════════════════════════════════════════════\n")
cat("Mean delta score (R2 − R1) for ESSENTIAL genes by stage and MOI\n")
cat("══════════════════════════════════════════════════════════════════\n")
cat("Negative delta = gene was being rescued by complementation in R1\n\n")

results_all %>%
  filter(essentiality == "Essential") %>%
  group_by(moi_r1, stage) %>%
  summarise(mean_delta = round(mean(delta), 3),
            sd_delta   = round(sd(delta),   3),
            .groups    = "drop") %>%
  pivot_wider(names_from = stage, values_from = c(mean_delta, sd_delta)) %>%
  print()

cat("\n══════════════════════════════════════════════════════════════════\n")
cat("Correlation: complementation fraction vs delta score\n")
cat("(essential genes only; tests if delta reveals complementation)\n")
cat("══════════════════════════════════════════════════════════════════\n")

results_all %>%
  filter(essentiality == "Essential") %>%
  group_by(moi_r1, stage) %>%
  summarise(
    r_comp_delta = round(cor(complementation, delta, method = "pearson"), 3),
    .groups      = "drop"
  ) %>%
  pivot_wider(names_from = stage, values_from = r_comp_delta) %>%
  print()
```



